A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-02-01 doc/large_strain.tex
Touch ups to mtrx-vec multiplication section; fix some spelling errors

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference? Martin's papers?}.
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
\changeJP{In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.}
}.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
{\color{red}
We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
{\color{red}as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless, }
the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
{\color{red}
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
{\color{red}as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.}

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
{\color{red}
In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
from unit to real cell \cite{kronbichler12}.
Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}

% likwid-mpirun -np 10 -nperdomain S:10 -g MEM_DP -m $EXC
{\color{red}
Finally, we analyze the node-level performance of the two approaches using the Roofline model.
Memory bandwidth (MBytes/s) and the number of floating point operations per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
The measurements are done on a single socket of RRZE cluster by pinning all 10 MPI processes to this socket, whose
memory bandwidth\footnote{As measured with the \textit{load\_avx} benchmark of \textit{likwid-bench} \cite{likwid}.} and peak performance are $B=47.2 \,\rm{GB/s}$ and $P=176\, \rm{GFlop/s}$, respectively.
Turbo mode of CPU was disabled.
}

@@

{\color{red}
Figure \ref{fig:roofline} shows results of the Roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_2d.pdf}
      \caption{Roofline (2D)}
      \label{fig:roofline_2d_tensor4}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_3d.pdf}
    \caption{Roofline (3D)}
    \label{fig:roofline_3d_tensor4}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_2d.pdf}
      \caption{Computing times (2D)}
      \label{fig:breakdown_stackedbar_2d_tensor4}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_3d.pdf}
    \caption{Computing times (3D)}
    \label{fig:breakdown_stackedbar_3d_tensor4}
  \end{subfigure}
  \caption{Breakdown of computing times and Roofline analysis of various steps of Algorithm \ref{alg:mf_tensor4}.}%
  \label{fig:breakdown}
\end{figure}

@@

In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
(i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
(ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
(iii) `quadrature loop' denotes operations performed on each quadrature (steps \mbox{6 -- 11}).
Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
global refinement in 2D and no global mesh refinement in 3D.
It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
`read / write' steps takes negligible amount of time spent on each element.
For higher order bases the majority of time is spent within the quadrature loop.
When analyzing the Roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
In this case the quadrature loop consists of, essentially, a
single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
Further optimization of \texttt{Tensor} classes in \textit{deal.II} are required to
improve the performance.
Such studies, however, are beyond the scope of the current work.
}

{\color{red}
To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order elements to two factors.
First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
bound, see Figure \ref{fig:roofline}.
Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
}

@@

{\color{red}
Additionally, we compare solution times of the preconditioned iterative solvers.
Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
On another hand, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engendering problems
with finite strain material models.
Extension of this study from the node level to the cluster level is beyond the scope of this work,
however we hypothesize that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library would provide a competitive solution for that case as well.}

@@

{\color{red}
The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth order and second order symmetric tensor
as well as single contraction of two second order tensors is performed.
Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}

{\color{red}
We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
}
with standard geometric transfer operations between each level.
The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
The numerical studies indicate that the proposed preconditioner
{\color{red}
leads to much fewer iterations of the iterative solver as compared to the algebraic multigrid preconditioner.
The multigrid matrix-free preconditioner also}
leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
Importantly the solution wall time per degree of freedom is close to being constant.
On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher order bases.
We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.
Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $FE^2$ homogenization
as well studying its behavior on the cluster level as opposed to node level, considered here.}

>>>>

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference? Martin's papers?}.
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
\changeJP{In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.}
}.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
{\color{red}
We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
{\color{red}as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{Arndt2017a} to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless, }
the matrix-free implementation is faster than the matrix-based counterpart already for \changeJP{bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D}.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
{\color{red}
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
{\color{red}as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.}

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point \changeJP{(``tensor4'')}, the matrix-free approach takes much less memory than its matrix-based counterpart.
{\color{red}
In those graphs we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
from unit to real cell \cite{kronbichler12}.
Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both \changeJP{deformed} and undeformed
configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}

% likwid-mpirun -np 10 -nperdomain S:10 -g MEM_DP -m $EXC
{\color{red}
Finally, we analyze the node-level performance of the two approaches using the roofline model.
Memory bandwidth (MBytes/s) and the number of floating point operations per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
The measurements are done on a single socket of RRZE cluster by pinning all 10 MPI processes to this socket, whose
memory bandwidth\footnote{As measured with the \textit{load\_avx} benchmark of \textit{likwid-bench} \cite{likwid}.} and peak performance are $B=47.2 \,\rm{GB/s}$ and $P=176\, \rm{GFlop/s}$, respectively.
Turbo mode of CPU was disabled.
}

@@

{\color{red}
Figure \ref{fig:roofline} shows results of the roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second-order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_2d.pdf}
      \caption{Roofline (2D)}
      \label{fig:roofline_2d_tensor4}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_3d.pdf}
    \caption{Roofline (3D)}
    \label{fig:roofline_3d_tensor4}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_2d.pdf}
      \caption{Computing times (2D)}
      \label{fig:breakdown_stackedbar_2d_tensor4}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_3d.pdf}
    \caption{Computing times (3D)}
    \label{fig:breakdown_stackedbar_3d_tensor4}
  \end{subfigure}
  \caption{Breakdown of computing times and roofline analysis of various steps of Algorithm \ref{alg:mf_tensor4}.}%
  \label{fig:breakdown}
\end{figure}

@@

In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
(i) `read / write' denotes reading and writing from/to a global vector (\changeJP{lines 2 and 13});
(ii) `sum factorization' denotes application of sum factorization techniques (\changeJP{lines} 4, 5 and 12);
(iii) `quadrature loop' denotes operations performed on each quadrature (\changeJP{lines \mbox{6 -- 11}}).
Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
global refinement in 2D and no global mesh refinement in 3D.
It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
`read / write' steps takes negligible amount of time spent on each element.
For higher-order bases the majority of time is spent within the quadrature loop.
When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
In this case the quadrature loop consists of, essentially, a
single contraction between two second-order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
double contraction between a symmetric second-order tensor and a fourth-order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
Further optimization of \texttt{Tensor} \changeJP{and \texttt{SymmetricTensor}} classes in \textit{deal.II} are required to
improve the performance.
Such studies, however, are beyond the scope of the current work.
}

{\color{red}
To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
bound, see Figure \ref{fig:roofline}.
Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
}

@@

{\color{red}
Additionally, we compare solution times of the preconditioned iterative solvers.
Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
\changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engendering problems
with finite strain material models.
Extension of this study from the node level to the cluster level is beyond the scope of this work,
however we hypothesize that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library would provide a competitive solution for that case as well.}

@@

{\color{red}
The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth-order and second-order symmetric tensor
as well as single contraction of two second-order tensors is performed.
Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}

{\color{red}
We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
}
with standard geometric transfer operations between each level.
The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
The numerical studies indicate that the proposed preconditioner
{\color{red}
leads to much fewer iterations of the iterative solver as compared to the algebraic multigrid preconditioner.
The multigrid matrix-free preconditioner also}
leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
Importantly the solution wall time per degree of freedom is close to being constant.
On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.
Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $FE^2$ homogenization
as well studying its behavior on the cluster level as opposed to node level, considered here.}
